The missing puzzle piece: Contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-05-26 manuscript/03-experiment.qmd
revision submitted

Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data. @fig-feature shows the strength of trend versus seasonality. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. Both measures are on a scale of [0,1]. the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @mstl.

@@

### Collaborative expert review to identify factors affecting consumption

To understand the external factors influencing product consumption, we began by visualizing the time series data for consumption as highlighted in the previous section. This initial step helped identify specific instances within the data that showed unusual patterns, such as peaks, unusually high or low observations, and consecutive periods of low consumption. These observations indicated potential events or factors affecting consumption levels.

To gather domain knowledge about factors affected the consumption of products, we contacted experts from EPSS who work in the distribution unit. Next, we collaborated with six experts, each with over 10 years of experience managing pharmaceutical distribution. In these sessions, we reviewed the time series of all products together to understand the potential drivers behind these anomalies. The experts provided insights based on their extensive experience and knowledge of distribution practices. When there was uncertainty or gaps in the data for certain products, additional experts were consulted to gather further context. If all experts reached a consensus, the influencing factors were confirmed and included in the analysis.

As part of this process, we examined the electronic bin cards for 33 pharmaceutical products and conducted detailed data exploration. A bin card is a physical or electronic record used to track the inventory levels of individual items in a storage location, such as a warehouse or stockroom. The term 'bin' refers to the specific storage location for an item. Information typically recorded on a bin card includes the Item ID, unit of measure, product description, account details, date, references, beginning balance, manufacturer, batch number, expiry date, quantity received, quantity issued, and balance. In summary, a bin card is a fundamental tool in inventory management, especially in environments where physical stock tracking is essential. It provides a straightforward and immediate way to monitor item quantities, helping to ensure effective inventory management. Over a period of two weeks, we meticulously traced consumption data through these bin cards. Moreover, We ensured accuracy by converting dates from the Ethiopian calendar to the Gregorian calendar and excluded transactions related to internal storage or warehouse relocation within EPSS. Through this collaborative approach, we were able to identify several predictors affecting distribution and consumption that are summarised as followings:

- Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.

<!-- - Stockouts:  occur when an inventory item is completely depleted, meaning there is no available stock to meet demand. This can happen when demand exceeds the available supply, or when there is a delay or failure in the replenishment process. Resulted in no transactions and a decline in consumption.  -->

<!-- - COVID-19: COVID-19, caused by the SARS-CoV-2 virus, had a profound impact on Ethiopia's pharmaceutical supply chain. The pandemic disrupted global manufacturing and supply, leading to shortages of essential medicines and raw materials, Ethiopia, like many other countries, relies heavily on imported pharmaceutical products. With the pandemic causing lockdowns and restrictions in manufacturing hubs like India and China, the supply of essential medicines and raw materials to Ethiopia was significantly disrupted. The impact of COVID-19 on consumption was significant. After the announcement of COVID-19 in Ethiopia, the central EPSS pushed a six-month supply of products to hubs to prevent stock outs. This push was from March to August 2020. During this period, if products were not stocked out, consumption continued to hubs and health facilities. -->

- Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced consumption. Stores closed during these periods, halting transactions. We observed increased consumption before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.

<!-- - Internal displacement: Occurs when people are forced to flee their homes but remain within their country's borders, typically due to conflicts, violence, or natural disasters. Conflicts, whether ethnic, political, or resource-driven, can lead to widespread instability, forcing large populations to move and creating humanitarian crises. Conflicts often lead to the destruction of infrastructure, including roads, bridges, and warehouses, making it difficult to transport pharmaceutical products to affected areas. This disruption can cause significant delays or prevent the delivery of essential medicines altogether. Internal displacements and conflicts in Ethiopia over the past five years also affected consumption and distribution. We traced periods of instability and included them as predictors. The identified conflict periods were September 2017 to March 2018, February 2019 to June 2019, and November 2020 to July 2022.  -->

- Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.

The key predictors identified are deterministic values, since both past and future values of these predictors are known in advance. We then integrated them into the consumption data, creating the complete dataset required for model building and running the experiment.

@@

@@

In all models, probabilistic forecasts were generated using bootstrapping to create 1,000 possible future scenarios for each period within the forecast horizon [@hyndman2021forecasting].

@@

To assess the performance of our forecasting methods, we split the data into a series of training and test sets and apply time series cross-validation with a forecast horizon of 6 months. Each training set is expanded in monthly increments, allowing model development and hyperparameter tuning to be performed on the training data, with errors evaluated on the corresponding test sets. We assess forecasting performance using both point and probabilistic error measures.

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution.
$$
  \text{CRPS} = \text{mean}(p_j),
$$ {#eq-CRPS}
where
$$
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
$$
where $G_j(x)$ is the forecasted probability distribution function for forecast horizon $j$, and $F_j(x)$ is the true probability distribution function for the same period.

@@

<!-- If the  $100(1−\alpha)\%$ prediction interval at time  t is given by $[\ell_{\alpha,t}, u_{\alpha,t}]$, then the Winkler score is defined as the length of the interval plus a penalty if the observation is outside the interval -->

<!-- $$ -->
<!-- W_{\alpha,t} = \begin{cases} -->
<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (\ell_{\alpha,t} - y_t) & \text{if } y_t < \ell_{\alpha,t} \\ -->
<!--   (u_{\alpha,t} - \ell_{\alpha,t})   & \text{if }  \ell_{\alpha,t} \le y_t \le u_{\alpha,t} \\ -->
<!--   (u_{\alpha,t} - \ell_{\alpha,t}) + \frac{2}{\alpha} (y_t - u_{\alpha,t}) & \text{if } y_t > u_{\alpha,t}. -->
<!--   \end{cases} -->
<!-- $$ -->

<!-- In this study, we report skill scores for winkler aand CRPS, with skill scores, we compute a forecast accuracy measure relative to some benchmark method. For example, if we use the naïve method as a benchmark, and also compute forecasts using the ETS model, we can compute the CRPS and Winkler skill score of the ETS model relative to the naïve method. If the skill scores are negative, it indicates that the proposed method performs worse than the naïve method; positive scores suggest it performs better. In this sense, skill scores are not entirely independent of data scale, similar to the concepts behind MASE and RMSSE for point forecasts. -->

<!-- $$ -->
<!-- \frac{\text{CRPS}_{\text{Naïve}} - \text{CRPS}_{\text{ETS}}}{\text{CRPS}_{\text{Naïve}}}. -->
<!-- $$ -->

<!-- $$ -->
<!-- \frac{\text{Winkler}_{\text{Naïve}} - \text{Winkler}_{\text{ETS}}}{\text{Winkler}_{\text{Naïve}}}. -->
<!-- $$ -->

>>>>

Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data. @fig-feature shows the strength of trend versus seasonality. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. Both measures are on a scale of [0,1]. the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @bandara2025mstl

@@

```{r}
#| label: tbl-feature
#| echo: false
#| cache: false
#| tbl-cap: "Summary statistics of the pharmaceutical product time series"
#| tbl-pos: "H"

library(fpp3)
all_feature <- med_qty_tsb %>%
  features(quantity, feature_set(pkgs = "feasts")) |> dplyr::select(`ACF lag 1`=acf1,zero_run_mean,nonzero_squared_cv)

summary_stat <- med_qty_tsb %>%
  features(quantity,list(Mean = mean, Median=median, `Standard deviation`=sd, Miinum=min, Maximum=max)) |> 
  bind_cols(all_feature) |> mutate(across(where(is.numeric), ~ round(.x, 2)))
summary_stat |> 
  knitr::kable(booktabs = T, linesep = "") |>
  kable_styling(latex_options = c("scale_down"))
```

@tbl-feature also provides the summary statistics computed for each pharmaceutical product time series include traditional distributional measures—mean, median, standard deviation, minimum, and maximum—alongside time series–specific and structural characteristics. These include the first-order autocorrelation (ACF at lag 1), the mean length of consecutive zero runs (zero_run_mean), and the squared coefficient of variation computed only on non-zero values (nonzero_squared_cv). Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero consumption.

## Collaborative expert review to identify factors affecting consumption

\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation. In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}

Figure \ref{fig:expert-pipeline} \textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}

```{=latex}
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.85, transform shape, node distance=1.3cm and 2cm]

% Node style
\tikzset{
  process/.style={
    rectangle,
    draw=black,
    thick,
    text width=4.5cm,
    align=center,
    minimum height=1.4cm
  },
  arrow/.style={
    thick,
    ->,
    >=Stealth
  }
}

% Nodes
\node (step1) [process] {1. Expert Identification (with EPSS)};
\node (step2) [process, below=of step1] {2. Onboarding \& Orientation};
\node (step3) [process, below=of step2] {3. Workshop: Collaborative Data Review \& Joint Identification of Anomalies};

\node (step4) [process, right=of step1] {4. Cross Validation Using Bin Cards \& Inventory Records};
\node (step5) [process, below=of step4] {5. Structured Expert Elicitation \& Contextual Clarification};

\node (step6) [process, right=of step4] {6. Expert Elicitation of Predictor Relevance};
\node (step7) [process, below=of step6] {7. Consensus-Based Selection of Predictors};
\node (step8) [process, below=of step7] {8. Variable Encoding \& Time Alignment};

\node (step9) [process, below=of step8, yshift=-0.3cm] {9. Integration into Forecasting Models};

% Arrows
\draw [arrow] (step1) -- (step2);
\draw [arrow] (step2) -- (step3);

\path (step3.east) -- +(0.6,0) coordinate (s3stub);
\draw [arrow] (step3.east) -- (s3stub) |- (step4.west);

\draw [arrow] (step4) -- (step5);

\path (step5.east) -- +(0.6,0) coordinate (s5stub);
\draw [arrow] (step5.east) -- (s5stub) |- (step6.west);

\draw [arrow] (step6) -- (step7);
\draw [arrow] (step7) -- (step8);
\draw [arrow] (step8) -- (step9);

% Group boxes
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step1)(step2)(step3), label={[black]90:Expert Engagement}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step4)(step5), label={[black]90:Validation}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step6)(step7)(step8), label={[black]90:Feature Construction}] {};
\node[draw=black, thick, dashed, inner sep=0.3cm, rounded corners, fit=(step9), label={[black]90:Model Integration}] {};

\end{tikzpicture}
\caption{Diagram illustrating the sequential steps involved in collecting, structuring, and utilizing domain knowledge for modeling.}
\label{fig:expert-pipeline}
\end{figure}

```

\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products. Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles). These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events. Dates were standardized to the Gregorian calendar.}

\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor. The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality. A further description of these variables are summarized as followings:}

- \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}

- \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced consumption. Stores closed during these periods, halting transactions. We observed increased consumption before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.}

- \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}

\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset. This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}

@@

\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training. Each LSTM model was trained independently for each product series. Hyperparameter tuning was performed in a preliminary phase using a subset of products. Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance. The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}

@@

## Generating probabilistic forecasts

\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption. Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction} [@wang2023].

\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals. Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions. Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}

\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors. The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate. To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates. This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}

\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts} [@hyndman2021forecasting].

\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}

@@

\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation} [@hyndman2021forecasting]. \textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods. In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021. The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon. At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles. After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data. This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available. This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance. Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration. No test set information was used during model selection or tuning.}

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution. For probabilistic evaluation, 1,000 future paths were simulated per series, enabling robust estimation of forecast uncertainty.

$$
  \text{CRPS} = \text{mean}(p_j),
$$ {#eq-CRPS}
where
$$
  p_j = \int_{-\infty}^{\infty} \left(G_j(x) - F_j(x)\right)^2dx,
$$
where $G_j(x)$ is the forecasted probability distribution function for forecast horizon $j$, and $F_j(x)$ is the true probability distribution function for the same period.

@@

### 2025-05-26 manuscript/04-results.qmd
revision submitted

The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin. We report the distribution of accuracy metrics across all rolling origins. This shows how models varies in providing accuracy across different origins. The y-axis displays models sorted by their MASE and RMSSE values, with the model exhibiting the lowest error positioned at the bottom. This model is the Ariam model, which incorporates exogenous variables. Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models. This underscores the critical importance of systematically collecting such expert-informed data alongside transactional consumption data.

@@

@fig-prob illustrates the forecast distribution accuracy measured by CRPS, which evaluates both forecasting calibration and interval sharpness. A smaller CRPS value indicates better overall performance. We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts. Notably, ARIMA achieved the highest probabilistic forecast accuracy, while LSTM models had the lowest. This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.

We observe that LSTM models exhibit a widespread error distribution compared to other models. This is attributed to their performance variability, which can significantly depend on the characteristics of demand patterns across different products. For example, the product "Amlodipine - 5mg - Tablet" demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero demand. Such patterns align well with the strengths of LSTM models, which excel at capturing long-term dependencies and managing complex temporal fluctuations.
In contrast, the demand for "Anti-Rho (D)" is erratic and sporadic, characterized by frequent, random fluctuations without any discernible long-term trends or seasonal patterns. This chaotic nature, combined with inconsistent and often low volumes, poses significant challenges for LSTM models. The lack of clear structure or stable patterns limits the models' ability to generalize effectively, resulting in poorer forecast performance. Although LSTM models can handle irregular data to some extent, they require sufficient and well-distributed training samples to achieve optimal performance. In our study, the majority of products exhibit erratic and sporadic demand patterns similar to "Anti-Rho (D)", resulting in this widespread error distribution.

<!-- In addition to reporting the forecast accuracy presented across different rolling origins, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-horizon. this is the accuracy reorted as the average across rolling origins and all products fro each given horizon. For both the point forecast and distributional accuracy  we can see that models incoporating information collected by domain experts improves compared to counterparts that that uses only the historical comsuptions.  -->

<!-- Overall, our results indicate that forecast produces using models with infromation colelcted via the domain expert engagement provides reliable forecasts and improves upon the models relying solely on historcial consuption.  -->

@@

Our analysis also highlights the value of using foundational time series forecasting models like TimeGpt, particularly in developing countries and the global health and humanitarian sectors. In these contexts, the lack of advanced analytical skills and systematic data collection can hinder effective forecasting and decision-making.

@@

In practice, point forecasts are commonly used despite their limitations, but they do not account for the inherent uncertainty associated with forecasts. The future is inherently uncertain, and effective planning requires considering alternative scenarios. Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different consumption levels may occur with varying probabilities. The primary purpose of probabilistic forecasting, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty. This figure displays the forecast distribution of consumption over a 6-month horizon using a density plot. For each month within the forecast period, a separate distribution is generated. The plot also includes the point forecast alongside 80% and 90% prediction intervals to illustrate potential variability.

@@

>>>>

The forecasting performance is reported in @fig-point, in which the average forecast accuracy over forecast horizon and across all products is calculated for each origin. We report the distribution of accuracy metrics across all rolling origins. This shows how models varies in providing accuracy across different origins. The y-axis displays models sorted by their MASE and RMSSE values, with the model exhibiting the lowest error positioned at the bottom. This model is the LSTM model, followed by ARIMA, which incorporate exogenous variables. Additionally, @fig-point indicates that predictors obtained through interactions with domain experts enhance point forecast accuracy across most models. This underscores the critical importance of systematically collecting such expert-informed data alongside transactional consumption data.

@@

@fig-prob illustrates the forecast distribution accuracy measured by CRPS, which evaluates both forecasting calibration and interval sharpness. A smaller CRPS value indicates better overall performance. We observed that incorporating domain knowledge improved forecast accuracy for most models, enhancing not only point forecasts but also probabilistic forecasts. Notably, LSTM and ARIMA yielded the most accurate probabilistic forecasts when identified predictors were incorporated. This aligns with the findings related to point forecast accuracy, reinforcing the earlier explanations for these results.

\textcolor{blue}{While LSTM models achieved the best overall forecast accuracy across products, their performance exhibited notable variability depending on the characteristics of individual demand patterns. For example, the product $Amlodipine - 5mg - Tablet$ demonstrates periods of extreme variability, with spikes in demand followed by periods of very low or zero consumption. Such patterns align well with the strengths of univariate LSTM models, which are adept at capturing long-term dependencies and managing complex temporal fluctuations. In contrast, the demand for $Anti-Rho (D)$ is erratic and sparse, with frequent random fluctuations and little structural consistency. This lack of clear temporal patterns can make it challenging for LSTM models to learn generalizable signals, particularly given the limited data length available for training. Although LSTM can manage irregular data to some extent, it performs best when patterns are consistent or cyclic. These variations across product types contributed to the observed distribution of forecast errors across the product portfolio. Moreover, although we expect that multivariate LSTM models benefit from expert-informed predictors, our results show that the univariate LSTM consistently achieved better forecast performance. This outcome may be attributed to the nature of the predictors—binary, static, or weakly aligned with short-term temporal dynamics—which can disrupt rather than enhance learning when added to a neural network sensitive to input configuration. Moreover, with relatively short historical series and limited training samples, the inclusion of additional variables may have led to overfitting or reduced generalization. These findings suggest that while LSTM models can effectively learn temporal patterns from consumption data alone, incorporating structured external knowledge requires careful feature engineering and alignment to be beneficial.}

@@

## Computational efficiency and resource considerations

\textcolor{blue}{In addition to forecast accuracy, it is also important to consider the computational efficiency of each model, particularly in settings where computational resources and technical expertise are limited.}

@tbl-runtime \textcolor{blue}{shows the total runtime required to train and generate forecasts across all 33 pharmaceutical product time series for each method for all origins. All models were implemented using R, except TimeGPT, which was run via Google Colab using a T4 GPU backend. All other models were executed on a local machine with 7 CPU cores} (11th Gen Intel(R) Core(TM) i5-1135G7 \@ 2.40 GHz and 8 GB RAM)

```{r}
#| label: tbl-runtime
#| echo: false
#| tbl-cap: "Computation time required for training and generating forecasts for each model across all products."
#| tbl-pos: "H"
runtime <- read.csv('../result/runtime_tbl.csv') |> 
  janitor::clean_names() |> 
  mutate(category = cut(run_time_in_seconds,
                        breaks = c(-Inf, 100, 1000, Inf), # Adjust the breaks
                        labels = c("low", "medium", "high"))) |> 
  rename(Model = model,
         'Runtime (Seconds)' = run_time_in_seconds,
         'Runtime type' = runtime_type)

runtime |> 
  knitr::kable(booktabs = T, linesep = "") |>
  kable_styling(latex_options = c("striped"), font_size = 11)
```

\textcolor{blue}{As shown in Figure 4 and 5, models that incorporated expert-informed regressors generally achieved better forecast accuracy compared to their univariate versions. This trend was most evident in classical models such as ARIMA and regression, where the inclusion of regressors led to a noticeable shift toward lower and more concentrated error distributions. However, this improvement came with an increase in computational cost. In all cases, the addition of regressors increased runtime—by 72\% in the regression model (from 27.48 to 47.38 seconds) and by 43\% in ARIMA (from 96.27 to 138.08 seconds). TimeGPT also showed a slight increase in runtime (from 2.57 to 3.72 seconds), although the total processing time remained extremely low overall. Moreover, LSTM exhibited the largest increase in runtime when regressors were added, rising from 6,892 to 7,234 seconds. This sharp increase reflects the sensitivity of neural networks to input configuration, especially in contexts with limited data, irregular demand, and noisy signals.}

\textcolor{blue}{These results underscore the importance of balancing model performance with implementation cost. While deep learning models such as LSTM offer strong potential when well-tuned, they demand significantly more computational resources and may be less robust when integrating static or weakly aligned contextual features. By contrast, TimeGPT, a foundational model pretrained on large-scale time series data, provides a compelling alternative. It required less than 4 seconds to forecast all products, offered competitive accuracy, and required no tuning or retraining—making it well-suited for practical use in low-resource environments.}

@@

In practice, point forecasts are commonly used despite their limitations, but they do not account for the inherent uncertainty associated with forecasts. The future is inherently uncertain, and effective planning requires considering alternative scenarios. Probabilistic forecasts offer a comprehensive approach by assigning likelihoods to a range of possible outcomes, recognizing that different consumption levels may occur with varying probabilities. \textcolor{blue}{The goal is to maximize the sharpness of these predictive distributions—i.e., how concentrated the forecasts are—while maintaining calibration, meaning that the predicted probabilities align well with actual outcomes} [@gneiting2014probabilistic]. \textcolor{blue}{This approach allows decision-makers to fully leverage available information, incorporating uncertainty in a structured and measurable way.} The primary purpose, as illustrated in @fig-forecast-density-hstep, is to quantify and communicate uncertainty. This figure displays the forecast distribution of consumption over a 6-month horizon using a density plot. For each month within the forecast period, a separate distribution is generated. The plot also includes the point forecast alongside 80% and 90% prediction intervals to illustrate potential variability.

Probabilistic forecasts enhance planning and decision-making by offering a comprehensive view of potential future outcomes and their likelihood, rather than relying on a single point estimate. A probabilistic forecast takes the form of a predictive probability distribution over future quantities or events of interest.

@@

\textcolor{blue}{To further illustrate the practical relevance, we consider an illustrative case where the mean forecasted consumption for the next month is 1,000 units, with a 90\% prediction interval ranging from 850 to 1,150 units.} @tbl-point_vs_probabilistic \textcolor{blue}{summarizes the decision outcomes under each approach.}

\textcolor{blue}{Under a traditional approach, inventory decisions would rely solely on the point forecast. Based on the mean prediction of 1,000 units, the store manager would order precisely that quantity, assuming it would meet the expected demand. However, this approach does not incorporate any adjustment for uncertainty, potentially leading to stockouts if actual demand exceeds 1,000 units, or excess inventory if consumption is lower.}

\textcolor{blue}{In contrast, utilizing the full probabilistic forecast allows the decision-maker to take variability into account. If a higher service level is required, for example 95\%, the inventory policy may be adjusted by stocking closer to the upper bound of the 90\% prediction interval (e.g., 1,150 units). If inventory holding costs are a major concern and the organization can tolerate a modest risk of shortage, the order quantity could remain closer to the median forecast of 1,000 units.}

```{r}
#| label: tbl-point_vs_probabilistic
#| tbl-pos: "H"
#| tbl-cap: "Comparison of ordering decisions under point and probabilistic forecasts - an illustrative example."

library(kableExtra)
df <- tibble(
  `Forecast Type` = c("Point Forecast Only", "Probabilistic Forecast (95% service level)"),
  `Ordering Quantity` = c("1,000 units", "1,150 units"),
  `Risk Consideration` = c(
    "No explicit consideration of uncertainty",
    "Adjusts inventory to account for demand variability"
  )
)

  df |> knitr::kable(booktabs = T, linesep = "") |>
  kable_styling(latex_options = c("scale_down"), font_size = 11)

\textcolor{blue}{This illustrative comparison indicates that point forecasts provide a single estimate without adjusting for forecast uncertainty, while probabilistic forecasts allow decision-makers to explicitly align inventory decisions with risk tolerance and service level requirements. This is particularly important for pharmaceutical products, where the consequences of stockouts or overstocking can be significant both operationally and clinically.}

## Managerial Implications

\textcolor{blue}{This study offers actionable lessons for supply chain managers, public health planners, and policymakers navigating the uncertainty of pharmaceutical demand—particularly in systems like Ethiopia, where operational constraints and demand volatility are the norm, not the exception.}

\textcolor{blue}{At present, national forecasting at the Ethiopian Pharmaceutical Supply Service (EPSS) still relies heavily on basic extrapolation tools—usually Excel sheets or donor-developed software such as the Quantification Analysis Tool (QAT). These tools are easy to use but limited: they assume stable demand, ignore uncertainty, and often miss the operational signals embedded in local experience. In practice, forecasts are sometimes adjusted based on gut feeling or anecdotal program insights—not because planners want to—but because the tools don’t offer a better alternative.}

\textcolor{blue}{The models developed here—freely available, open-source, and designed to integrate directly with routine EPSS data—allow planners to make decisions using full probabilistic forecasts, not just single-point estimates. Forecasts are delivered with prediction intervals (e.g., 80\% or 90\%) that help teams decide not only how much to order, but how much risk they’re willing to tolerate. For products with known seasonality—like antimalarials or diagnostic kits—this matters. The system doesn’t just forecast a number; it gives planners a buffer strategy.}

\textcolor{blue}{Beyond accuracy, the models also reflect how medicines actually move through the system. Predictors based on warehouse replenishments, fiscal year inventory routines, and known disease cycles are embedded into the forecasts—not hard-coded, but learned from the data in ways that are transparent and reproducible. These inputs are drawn from expert operational knowledge and can be updated or modified as the system evolves.}

\textcolor{blue}{Importantly, the entire modeling pipeline is built in R and Python, and is already shared alongside code and data. This makes it immediately usable by EPSS analysts and regional logistics teams, even in settings without access to proprietary tools or high-end infrastructure. It's not a black box; it's something the system can own, modify, and grow with.}

\textcolor{blue}{To support real uptake, we recommend five practical steps:}

- \textcolor{blue}{Embed these probabilistic models into EPSS quantification rounds to replace rigid extrapolation methods and bring scenario-based planning into monthly and quarterly cycles.}

- \textcolor{blue}{Use prediction intervals to define risk-informed buffer policies, especially for high-volatility products. Not every item needs the same margin of error.}

- \textcolor{blue}{Automate the collection of domain knowledge  (e.g., replenishment events, stock counts, seasonal triggers) into the logistics data stream to reduce reliance on manual elicitation while preserving contextual intelligence.}

- \textcolor{blue}{Train regional and central teams using the open-source tools and shared codebase, turning forecasting into a hands-on, in-country competency rather than a dependency on external tools or consultants.}

- \textcolor{blue}{Establish routine forecast review sessions, supported by visual dashboards that reflect not just what the model says, but how uncertain that estimate is—and what that means for stock levels, procurement plans, and emergency readiness.}

\textcolor{blue}{Taken together, these steps push forecasting beyond formulas and into real decision-making.
}
